---
title: "Exploratory Data Analysis"
editor: visual
---

## Setup

```{r}
# Save packages as a vector
all.lib<-c("tidyverse","ggplot2", "dplyr","tidyr","modelr")

# Load packages
lapply(all.lib,require,character.only=TRUE)
```

## Introduction:

Steps to exploratory data analysis:

1.  Generate questions and hypothesis about the data.
2.  Understand your data
3.  Read the metadata if the data is not yours
4.  Think about the analysis plan led by questions
5.  Make sure your hypothesis-driven studies are clearly stated

## Load your data: Example 1

a\. Fuel economy data from 1999 to 2008 for 38 popular models of cars. The subset dataset from EPA. Full data available \[here\](https://fueleconomy.gov)

```{r}
# Load data
data("mpg")
mpg
```

## Exploring the data 

```{r}
# Useful functions in exploring the data
ncol(mpg)
nrow(mpg)
dim(mpg)
str(mpg)
summary(mpg)
head(mpg)
tail(mpg)
```

## Load your data: Example 2

b\. Flow of the River Nile, measurements of the annual flow of the river Nile at Aswan from Balke, N. S. (1993) and Cobb, G. W. (1978)

## Another example of time series data

```{r}
data("Nile")
head(Nile)
table(Nile)
```

## Creating a tibble

```{r}
data <- data.frame(a = 1:3, b = letters[1:3], c = Sys.Date() - 1:3)

data

#>   a b          c
#> 1 1 a 2023-02-21
#> 2 2 b 2023-02-20
#> 3 3 c 2023-02-19

as_tibble(data)

#> # A tibble: 3 × 3
#>       a b     c         
#>   <int> <chr> <date>    
#> 1     1 a     2023-02-21
#> 2     2 b     2023-02-20
#> 3     3 c     2023-02-19

## Creating a tibble from preloaded dataset
data(iris)
head(iris)
as_tibble(iris)

# Saving the tibble a new dataset
iris.updated<-as_tibble(iris)
```

## Finding patterns in data through visualization and transformations

### Categorical variable

**Using geom_bar for displaying categorical variable,**

```{r}
#| label: fig-onefig
#| fig-cap: "Displaying categorical variable with mpg."
#| fig-align: "center"
ggplot(data = mpg) +
  geom_bar(mapping = aes(x = manufacturer))
```

**Using the same above example for iris dataset, using the categorical variable,**

```{r}
#| label: fig-twofig
#| fig-cap: "Displaying categorical variable with iris."
#| fig-align: "center"
ggplot(data = iris) +
  geom_bar(mapping = aes(x = Species))
```

Notice how the bars are equal because the y axis that shows the count is same (i.e 50) for each of the species.

### Count function

The height of the bars displays how many observations occurred with each x value.

```{r}
mpg %>% 
  count(manufacturer)
```

```{r}
iris.updated %>% 
  count(Species)
```

### Continuous variable

**Using geom_bar for a continuous variable,**

```{r}
#| label: fig-threefig
#| fig-cap: "Displaying continuous variable with iris."
#| fig-align: "center"
ggplot(data = iris.updated) +
  geom_bar(mapping = aes(x = Sepal.Length))
```

**Using histogram is better here for continuous variable,**

```{r}
#| label: fig-fourthfig
#| fig-cap: "Displaying continuous variable using histogram, with iris."
#| fig-align: "center"
iris.updated %>%
  ggplot()+
  geom_histogram(mapping = aes(x = Sepal.Length),binwidth= 0.1)
```

**Using mpg dataset,**

```{r}
#| label: fig-fivefig
#| fig-cap: "Displaying continuous variable with mpg."
#| fig-align: "center"
mpg %>%
 ggplot()+
 geom_histogram(mapping = aes(x = displ), binwidth = 0.5)
```

### cut_width function

Display how the histogram was made

**For iris dataset,**

```{r}
iris.updated %>% 
count(cut_width(Sepal.Length, 0.1))
```

**For mpg dataset,**

```{r}
mpg %>% 
count(cut_width(displ, 0.5))
```

-   Note here that the numbers correspond to the binwidth used in each histogram

### Filter function

smaller dataset by filtering. The filter function evaluates a condition inside the bracket.

```{r}
mpg %>% 
  filter(cty >20)

mpg.sub<- mpg %>%
  filter(cty >20)
```

```{r}
iris.updated %>%
  filter(Petal.Width > 1)

iris.sub<- iris.updated %>%
  filter(Petal.Width >1)
```

**Plotting a subset of the dataset,**

```{r}
#| label: fig-sixfig
#| fig-cap: "Plotting a subset dataset, mpg."
#| fig-align: "center"
# plotting 
ggplot(data = mpg.sub, mapping = aes(x = cty)) +
  geom_histogram(binwidth = 0.25)
```

```{r}
#| label: fig-sevenfig
#| fig-cap: "Plotting a subset dataset, iris."
#| fig-align: "center"
#  plotting 
ggplot(data = iris.sub, mapping = aes(x = Petal.Width)) +
  geom_histogram(binwidth = 0.1)
```

### Expanding the x-y limits

```{r}
#| label: fig-eightfig
#| fig-cap: "Expanding the x-y coordinates,mpg."
#| fig-align: "center"
# Seeing unusual values
ggplot(mpg) + 
  geom_histogram(mapping = aes(x = displ), binwidth = 0.5)+
  coord_cartesian(ylim = c(0, 50), xlim=c(0,100))
```

### Select function

Used for subsetting a dataset, by selecting column names,

```{r}
mpg.filtered1 <- mpg %>% 
  filter(cty < 10 | cty > 15) %>% 
  select(manufacturer,cty, displ)

mpg.filtered1
```

### Arrange function

Sorts in ascending order by the column name provided,

```{r}
mpg.filtered2.asc <- mpg %>% 
  filter(cty < 10 | cty > 15) %>% 
  select(manufacturer,cty, displ) %>%
  arrange(cty)

mpg.filtered2.asc
```

Using arrange to sort in descending order, use the function across within arrange.

-   Note remember you can nest the functions within another function.

    ```{r}
    mpg.filtered2.dsc <- mpg %>% 
      filter(cty < 10 | cty > 15) %>% 
      select(manufacturer,cty, displ) %>%
      arrange(across(cty,desc))

    mpg.filtered2.dsc
    ```

### Mutate function

Used for adding more columns using a condition. Note here that the notation for

```         
 ifelse(CONDITION, VALUE_IF_TRUE, VALUE_IF_FALSE)
```

```{r}
# Mutate - add more columns
mpg.filtered3 <- mpg.filtered2.asc %>% 
  mutate(newcol = ifelse(cty < 10, NA, 0))

mpg.filtered3
```

**You can add multiple columns using mutate function,**

```{r}
iris.updated %>%
  mutate(order.name="Asparagales",
         height=Sepal.Length*5)
```

### Boxplot

Using function geom_boxplot to look at trends in data,

```{r}
#| label: fig-ninefig
#| fig-cap: "Using boxplots, mpg."
#| fig-align: "center"
# Boxplot
ggplot(data = mpg, mapping = aes(x = manufacturer, y = hwy)) +
  geom_boxplot()+
  xlab("highway miles")+
  ylab("manufacturer")
```

### Reorder function

-   Using function reorder to see a clear trend.
-   Note that reorder is a special case of the function factor.
-   The factor function is in base R and is used for ordering vector data.

```{r}
#| label: fig-tenfig
#| fig-cap: "Clear trend using reorder function."
#| fig-align: "center"
# Easy to see trend
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(manufacturer, hwy, FUN = median), y = hwy))+
  xlab("highway miles")+
  ylab("manufacturer")
```

### Flip the coordinates

```{r}
#| label: fig-elevenfig
#| fig-cap: "Flipped coordinates,mpg."
#| fig-align: "center"
# Flip coordinates
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(manufacturer, hwy, FUN = median), y = hwy))+
  coord_flip()+
  xlab("highway miles")+
  ylab("manufacturer")
```

### Scatterplot of the data

Geom point is used for looking at a scatterplot,

```{r}
#| label: fig-twelvefig
#| fig-cap: "Scatterplot of the data,mpg."
#| fig-align: "center"
# Scatterplot
ggplot(data = mpg) +
  geom_point(mapping = aes(x = manufacturer, y = hwy))+
   xlab("highway miles")+
   ylab("manufacturer")
```

## Finding patterns in data through modelling

Refine the questions based on what you learn and repeat the process. Let's ask a new question.

*How highway miles are varying with city miles?*

```{r}
#| label: fig-thirteenfig
#| fig-cap: "Scatterplot of city miles v/s highway miles."
#| fig-align: "center"
# How highway miles are varying with city miles
ggplot(data = mpg)+ 
  geom_point(mapping = aes(x = hwy,y=cty))+
   xlab("highway miles")+
   ylab("city miles")
```

### Linear regression

*Are highway miles linearly related with city miles?*

### Defining a linear model

```{r}
lm(hwy~cty,data = mpg) 
model1<-lm(hwy~cty,data = mpg)
summary(model1)
plot(model1)

# Other useful functions
coefficients(model1) # model coefficients
confint(model1, level=0.95) # Confidence Intervals for model parameters

residuals(model1) # residuals values
plot(residuals(model1)) # residuals plots
```

### Using add_residual function within tidyverse

```{r}
#| label: fig-fourteenfig
#| fig-cap: "Adding residuals to ggplot,mpg."
#| fig-align: "center"
# Using ggplot2 and add_residual function
mpg.model1 <- mpg %>% 
  add_residuals(model1) %>% 
  mutate(resid = exp(resid))

ggplot(data = mpg.model1) + 
  geom_point(mapping = aes(x = hwy, y = resid))
```

### Is there a logarithmic relationship? 

```{r}
lm(log(hwy)~cty,data = mpg)
```

### Finding correlation

-   cor() : computes the correlation coefficient
-   cor.test() : test for association/correlation between paired samples. It returns both the correlation coefficient and the significance level(or p-value) of the correlation.

```{r}
# Default method: Pearson
cor(mpg$cty, mpg$hwy)
cor.test(mpg$cty, mpg$hwy, method="pearson")

# Save into a vector
cor.result<-cor.test(mpg$cty, mpg$hwy, method="pearson")
```

-   t is the t-test statistic value;
-   df is the degrees of freedom;
-   p-value is the significance level of the t-test;
-   conf.int is the confidence interval of the correlation coefficient at 95%;
-   sample estimates is the correlation coefficient(Cor.coeff)

```{r}
# Extract the p.value
cor.result$p.value

# Extract the correlation coefficient
cor.result$estimate
```

### T-test

```{r}
t.test(mpg$cty, mpg$hwy) # here both the variable are numeric

# paired t-test
t.test(mpg$cty,mpg$hwy,paired=TRUE) # both the variable are numeric
```

i.  You can use the var.equal = TRUE option to specify equal variances and a pooled variance estimate.

ii. You can use the alternative="less" or alternative="greater" option to specify a one tailed test.

## Simulated dataset using probability distribution

```{r}
# Creating simulated data to work
set.seed(5)

xvar <- 1:20 + rnorm(20,sd=3)
zvar <- (1:20)/4 + rnorm(20,sd=2)
yvar <- -2*xvar + xvar*zvar/5 + 3 + rnorm(20,sd=4)

# Make a data frame 
mydat <- data.frame(x=xvar, y=yvar, z=zvar)

# first 6 rows
head(mydat)
```

### T-test

```{r}
# independent 2-group t-test
#t.test(y~x) # where y is numeric and x is a binary factor

# independent 2-group t-test
t.test(yvar,xvar) # where y1 and y2 are numeric

# paired t-test
t.test(yvar,xvar,paired=TRUE) # where yvar,xvar are numeric
```

### Correlation

#### Pearson's

```{r}
# Correlation between xvar and yvar 
# Default method: Pearson
cor(mydat$x, mydat$y)
cor.test(mydat$x, mydat$y, method="pearson")

# Save into a vector
cor.result<-cor.test(mydat$x, mydat$y, method="pearson")

# Extract the p.value
cor.result$p.value

# Extract the correlation coefficient
cor.result$estimate
```

#### Spearman's

```{r}
# Spearman's correlation
cor(mydat$x, mydat$y, method="spearman")
cor.test(mydat$x, mydat$y, method="spearman")
```

#### Kendall's 

```{r}
# Kendall's correlation
cor(mydat$x, mydat$y, method="kendall")
cor.test(mydat$x, mydat$y, method="kendall")
```
